---
title: "Week 4 Exam"
output:
  html_document: default
  pdf_document: default
---

***Preparation***
\
Let's first plan out how the data flows will work by identifying what packages, functions and data we need.

*Packages to load*
\ 
  - sf\
  - dplyr\
  - tidyverse\
  - here\
  - readxl\
  - janitor\
  - stringr\
  - tmap\

*Functions*
\ 
  - read_xlsx\
  - st_layers\
  - filter\
  - read.csv\
  - mutate\
  - pivot\
  - cleannames\

*Data*
\
Gender Inequality Index for 2010 and 2019\
\
World Dataset  - <https://services.arcgis.com/P3ePLMYs2RVChkJx/arcgis/rest/services/World_Countries_(Generalized)/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson>
\
Also, we can plan what our basic process will be and what the expected outputs are:

*Process*
\
  - load packages\
  - read data\
  - clean names\
  - pivot wider\
  - mutate column to calculate difference between indexes\
  - show difference on map\
  - save output\

*Output*
\
  - Shapefile of differences\
  - CSV of differences\

***Code***

By following our plan above we start by loading the required packages.

```{r}
library(sf)
library(dplyr)
library(tidyverse)
library(here)
library(readxl)
library(janitor)
library(stringr)
library(tmap)
```

Next we will read in the required data.  In this case the Gender Inequality Index data whihc was downloaded from https://hdr.undp.org/data-center/documentation-and-downloads and also the world geometries geojson.

```{r}
# read gender data and prepare the relevant columns
gii_raw <- readxl::read_xlsx(here::here("data", "GenderInequalityIndex2010_2019.xlsx")) %>%
  janitor::clean_names() %>%
  dplyr::select(country_iso_code,country, year, value)

# read geometries and prepare columns
world <- sf::st_read("https://services.arcgis.com/P3ePLMYs2RVChkJx/arcgis/rest/services/World_Countries_(Generalized)/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson") %>%
  janitor::clean_names() %>%
  dplyr::select(country, iso, geometry)
```

Now we need to prepare the gender data, firstly by removing any regions from the GII country column.

```{r}
# create data frame and filter to only countries (exclude regions)  
gii_filtered <- data.frame(gii_raw) %>%
  filter(str_detect(gii_raw$country_iso_code,"ZZ",TRUE))
```

We now pivot the data so it is in a usable format, rename the columns and create a new column that calculates the difference between the 2019 and 2010 index values.

```{r}
# pivot gii data
gii_wider <- gii_filtered %>%
  pivot_wider(names_from = "year",values_from = "value")

#rename value columns for ease of use
gii_wider <- gii_wider %>% dplyr::rename(value2010 = "2010", value2019 = "2019")

# create new column showing difference between 2019 and 2010 values
gii_new <- gii_wider %>%
  mutate(., difference=(value2019-value2010))
```

Then spatialise the data by joining the GII data with the world geometries geojson that was read in earlier.

```{r}
# spatialise the data by joining to world file
world_gii_spatial <- world %>%
  left_join(., 
            gii_new,
            by = c("country"="country"))
```

We can view the data on a map at this stage usibg tmap.

```{r}
# view the data on a map
tmap_mode("plot")
tm_shape(world_gii_spatial) + 
  tm_polygons("difference", 
              style="jenks",
              palette="Greens",
              midpoint=NA,
              title="GII difference",
              alpha = 0.5) + 
  tm_basemap(server = "OpenStreetMap") +
  tm_layout(title = "Difference between Gender Inequlaity Index between 2010 and 2019")
```

The final stage is to write the data to a CSV and a shapefile.

```{r}
# write data csv and shapefile
write.csv(gii_new, file = here::here("output", "GenderInequalityIndexWorld.csv"))
st_write(world_gii_spatial, "output/GenderInequalityIndexMap.shp", append=FALSE)
```


